---
title: "Reproducible Research: Course Project 1"
author: "Rushil Bala"
output: html_document
---

## Loading and preprocessing the data

* Appropriately set the working directory.
* Download the raw data.

```{r results = "hide", warning = FALSE, message = FALSE}
library(ggplot2)
library(dplyr)
library(Hmisc)
```

```{r}
if(!file.exists("activity.csv")){
  unzip("activity.zip")
}
rawdata <- read.csv("activity.csv", header = TRUE)
```

## What is mean total number of steps taken per day?

```{r}
total_steps <- tapply(rawdata$steps, rawdata$date, FUN = sum, na.rm = TRUE)
qplot(total_steps, binwidth = 500, fill = "red", xlab = "Total number of steps moved every day", ylab = "Frequency") +
  theme(legend.position = "none")
```

* Mean: `r mean(total_steps, na.rm = TRUE)`
* Median: `r median(total_steps, na.rm = TRUE)`

## What is the average daily activity pattern?

```{r}
average_steps <- aggregate(x = list(avg_steps = rawdata$steps), by = list(fm_interval = rawdata$interval), FUN = mean, na.rm = TRUE)
ggplot(data = average_steps, aes(x = fm_interval, y = avg_steps, col = "red")) +
  geom_line() +
  labs(x = "5-minute intervals", y = "Average number of steps moved every day") +
  theme(legend.position = "none")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
ans <- average_steps[which.max(average_steps$avg_steps), 1]
```

* Answer: `r ans`

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as `NA`. The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r}
num_NA <- length(which(is.na(rawdata$steps)))
```

* Number of NA: `r num_NA`

```{r}
imputeddata <- rawdata
imputeddata$steps <- impute(rawdata$steps, FUN = mean)
imputeddatadate <- tapply(imputeddata$steps, imputeddata$date, sum)
qplot(imputeddatadate, binwidth = 500, fill = "red", xlab = "Total number of steps moved every day", ylab = "Frequency") +
  theme(legend.position = "none")
```

* Mean: `r mean(imputeddatadate)`
* Median: `r median(imputeddatadate)`

Note that `NA` values were initially ignored (`na.rm = TRUE`). The mean and median increased after these were replaced by the initial means.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
imputeddata$datecat <- ifelse(weekdays(as.Date(imputeddata$date)) == "Saturday" | weekdays(as.Date(imputeddata$date)) == "Sunday", "Weekend", "Weekday")
average_imputeddata <- aggregate(data = imputeddata, steps ~ interval + datecat, FUN = mean)
ggplot(average_imputeddata, aes(x = interval, y = steps, col = "red")) +
  geom_line() +
  facet_grid(datecat ~ .) +
  labs(x = "5-minute intervals", y = "Average number of steps moved") +
  theme(legend.position = "none")
```
